Bridging Gaps: Investigating COVID-19's Influence on Health Disparities in Connecticut

### 2023-12-15 manuscript/manuscript.tex
added descriptions of social need variables in introduction

%% preamble: Keep it clean; only include those you need
\usepackage{amsmath}
\usepackage[margin = 1in]{geometry}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{natbib}
% for space filling
\usepackage{lipsum}
% highlighting hyper links
\usepackage[colorlinks=true, citecolor=blue]{hyperref}
\usepackage{tabularx}
\usepackage{adjustbox}
\usepackage[utf8]{inputenc}

@@

\section{Introduction}\label{sec:intro}

Current research focuses on how social determinants of health (SDoH) plays a  massive
impact on one's health; it is estimated that 80 percent of a population's health outcomes are 
dictated by SDoH \citep{HOOD2016129}. Often, SDoH, when referring to an individual, can result in racial 
disparities in care when looking at a population\citep{Monroe2023-uq}. It has been shown that major inefficiencies
in the health system are attributed to overlooked prevention opportunities and unequal access
to care.\citep{Allin2014-xn}

@@

Despite the growing body of research on SDoH, there is a notable gap concerning the specific impact of COVID-19 on 
these determinants within diverse communities. This research aims to bridge this gap by comprehensively assessing how 
the pandemic has influenced key social determinants of health across various counties and racial groups in Connecticut. 
By identifying the specific ways in which different communities were affected, this study contributes valuable insights 
for targeted interventions, policy-making, and the development of equitable healthcare strategies.

@@

- The data will be presented in Section \ref{sec:data}

- The methods are described in Section \ref{sec:meth}

- The results are reported in Section \ref{sec:resu}

- A discussion concludes in Section \ref{sec:disc}

@@

In this study, descriptive statistics is utilized to outline the total population, racial composition, 
education levels, and average rehospitalization rate across the 8 counties in Connecticut. Chi-squared analyses were 
conducted to assess the significance of the difference of the variables of median income, poverty level, health insurance, 
utilities access, and electronics access between counties and across the four years. Additional 
Tukey's HSD tests were conducted to determine the specific counties and years that have had significant differences within 
each of the variables for each county.

@@

\ref{tab:Demographics1}
\ref{tab:Demographics2}

\ref{fig:Education by County Graph}
discuss, use lit to support why

\ref{fig:Social Needs by County Graph}
differences by year

\ref{tab:Multiple Comparison for Income}
the specific values for income

\ref{tab: Multiple Comparisons}
what discussion talks about

@@

@@

    Fairfield County & Hartford County & 20565.50* & 1883.47 & \textbf{\textless 0.001} & 14327.60 & 26803.40 \\
    & Litchfield County & 14919.50* & 1883.47 & \textbf{\textless 0.001} & 8681.60 & 21157.40 \\
    & Middlesex County & 9671.75* & 1883.47 & \textbf{<.001} & 3433.85 & 15909.65 \\
    & New Haven County & 25662.75* & 1883.47 & \textbf{<.001} & 19424.85 & 31900.65 \\
    & New London County & 21456.50* & 1883.47 & \textbf{<.001} & 15218.60 & 27694.40 \\

@@

    & Windham County & 28671.00* & 1883.47 & \textbf{<.001} & 22433.10 & 34908.90 \\

@@

    Hartford County & Fairfield County & -20565.50* & 1883.47 & \textbf{<.001} & -26803.40 & -14327.60 \\

@@

    & Middlesex County & -10893.75* & 1883.47 & \textbf{<.001} & -17131.65 & -4655.85 \\

@@

    & Tolland County & -11860.50* & 1883.47 & \textbf{<.001} & -18098.40 & -5622.60 \\

@@

    Litchfield County & Fairfield County & -14919.50* & 1883.47 & \textbf{<.001} & -21157.40 & -8681.60 \\

@@

    & New Haven County & 10743.25* & 1883.47 & \textbf{<.001} & 4505.35 & 16981.15 \\

@@

    & Windham County & 13751.50* & 1883.47 & \textbf{<.001} & 7513.60 & 19989.40 \\

@@

    Middlesex County & Fairfield County & -9671.75* & 1883.47 & \textbf{<.001} & -15909.65 & -3433.85 \\
    & Hartford County & 10893.75* & 1883.47 & \textbf{<.001} & 4655.85 & 17131.65 \\

@@

    & New Haven County & 15991.00* & 1883.47 & \textbf{<.001} & 9753.10 & 22228.90 \\
    & New London County & 11784.75* & 1883.47 & \textbf{<.001} & 5546.85 & 18022.65 \\

@@

    & Windham County & 18999.25* & 1883.47 & \textbf{<.001} & 12761.35 & 25237.15 \\

@@

    New Haven County & Fairfield County & -25662.75* & 1883.47 & <.001 & -31900.65 & -19424.85 \\

@@

    & Litchfield County & -10743.25* & 1883.47 & <.001 & -16981.15 & -4505.35 \\
    & Middlesex County & -15991.00* & 1883.47 & <.001 & -22228.90 & -9753.10 \\

@@

    & Tolland County & -16957.75* & 1883.47 & <.001 & -23195.65 & -10719.85 \\

@@

    New London County & Fairfield County & -21456.50* & 1883.47 & <.001 & -27694.40 & -15218.60 \\

@@

    & Middlesex County & -11784.75* & 1883.47 & <.001 & -18022.65 & -5546.85 \\

@@

    & Tolland County & -12751.50* & 1883.47 & <.001 & -18989.40 & -6513.60 \\

@@

    & Hartford County & 11860.50* & 1883.47 & <.001 & 5622.60 & 18098.40 \\

@@

    & New Haven County & 16957.75* & 1883.47 & <.001 & 10719.85 & 23195.65 \\
    & New London County & 12751.50* & 1883.47 & <.001 & 6513.60 & 18989.40 \\
    & Windham County & 19966.00* & 1883.47 & <.001 & 13728.10 & 26203.90 \\

@@

    & Litchfield County & -13751.50* & 1883.47 & <.001 & -19989.40 & -7513.60 \\
    & Middlesex County & -18999.25* & 1883.47 & <.001 & -25237.15 & -12761.35 \\

@@

    & Tolland County & -19966.00* & 1883.47 & <.001 & -26203.90 & -13728.10 \\

@@

@@

\begin{table}[]
  \label{tab: Multiple Comparisons}
  \caption{Multiple Comparisons by Year and Town}
  \resizebox{\columnwidth}{!}{%
  \begin{tabular}{llrrrrl}
  \textbf{Social Need}   & \textbf{Comparison by} & \multicolumn{1}{l}{\textbf{DF}} & \multicolumn{1}{l}{\textbf{Sum of Squares}} & \multicolumn{1}{l}{\textbf{Mean Square}} & \multicolumn{1}{l}{\textbf{F Value}} & \textbf{P Value}           \\
  Median Income          & Year                   & 3                               & 158657606                                   & 52885869                                 & 95.57                                & <.0001                     \\
                         & Town                   & 7                               & 2618521106                                  & 374074444                                & 675.99                               & <.0001                     \\
  Food                   & Year                   & 3                               & 3.6450625                                   & 1.2150208                                & 10.08                                & 0.0003 \\
                         & Town                   & 7                               & 348.3382875                                 & 49.7626125                               & 412.72                               & <.0001                     \\
  Housing                & Year                   & 3                               & 1.4535125                                   & 0.4845042                                & 0.41                                 & 0.7506 \\
                         & Town                   & 7                               & 183.9498375                                 & 26.2785482                               & 21.99                                & <.0001                     \\
  Rehospitalization Rate & Year                   & 3                               & 0.00050938                                  & 0.00016979                               & 1.45                                 & 0.2578 \\
                         & Town                   & 7                               & 0.00427188                                  & 0.00061027                               & 5.2                                  & 0.0015 \\
  Uninsured Rate         & Year                   & 3                               & 2.6848375                                   & 0.89494583                               & 17.17                                & <.0001                     \\
                         & Town                   & 7                               & 92.8554875                                  & 13.26506964                              & 254.57                               & <.0001                    
  \end{tabular}%
  }
  \end{table}

@@

The results of the analysis shed light on the intricate relationships between county, race, and key 
variables such as median income, housing affordability, rehospitalization rates, food stamps usage, 
uninsured rates, and racial demographics. Understanding these correlations is vital in comprehending 
the socio-economic and racial disparities prevalent in the studied region.

The disparity in median income over the studied years, particularly the peak observed in 2020, 
signifies economic fluctuations. Notably, county 1 stands out with significantly higher income 
compared to other counties, indicating potential disparities in resource access and opportunities. 
This wealth gap implies potential repercussions on community health outcomes. The consistent trend in 
the percentage of income spent on rent underscores a persistent challenge for residents, especially 
noticeable in counties 1 and 5, indicating enduring economic pressure.

@@

The racial demographics highlight disparities in various counties. County 3's higher percentage of white 
individuals, coupled with County 2, 5, and 1 having the highest black population, points to the need for
targeted interventions addressing racial health inequalities. These disparities might be indicative of 
historical, social, and economic factors influencing the healthcare experiences of different racial groups.

>>>>

%% preamble: Keep it clean; only include those you need
\usepackage{amsmath}
\usepackage[margin = 1in]{geometry}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{natbib}
% for space filling
\usepackage{lipsum}
% highlighting hyper links
\usepackage[colorlinks=true, citecolor=blue]{hyperref}
\usepackage{tabularx}
\usepackage{adjustbox}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{threeparttable}

@@

\section{Introduction}\label{sec:intro}
\\
I. Prevalence of Social Determinants of Health 
\\
Current research focuses on how social determinants of health (SDoH) plays a  massive
impact on one's health; it is estimated that 80 percent of a population's health outcomes are 
dictated by SDoH \citep{HOOD2016129}. Often, SDoH, when referring to an individual, can result in racial 
disparities in care when looking at a population\citep{Monroe2023-uq}. It has been shown that major inefficiencies
in the health system are attributed to overlooked prevention opportunities and unequal access
to care.\citep{Allin2014-xn}

@@

Despite the growing body of research on SDoH, there is a notable gap concerning the specific impact of COVID-19 on 
these determinants within diverse communities. This research aims to bridge this gap by comprehensively assessing how 
the pandemic has influenced key social determinants of health across the 8 CT counties and a four year timeframe prior to 
the start of the COVID pandemic. 
By identifying the specific ways in which different communities were affected, this study contributes valuable insights 
for targeted interventions, policy-making, and the development of equitable healthcare strategies.
\\
II. Introduction to the Impact of Social Needs(specify for CT)
\\
This study will compare each of the following social needs by Year(2017-2020) and by 8 CT Counties: 
Median Income,
Educaation,
Food Access
Housing Stability,
Health Insurance,
Reshospitalization rates.

Income is a fundamental
social need that significantly influences
one's health. The level of income
directly affects an individual's access
to essential resources such as nutritious
food, healthcare, and secure housing.
People with lower incomes often face
barriers to healthcare services and
struggle to afford a balanced diet,
leading to higher risks of chronic
illnesses. Income disparities contribute
to inequalities in health outcomes, as
those with limited financial resources
may experience higher stress levels and
reduced access to preventive care.
Addressing income inequality is crucial
for promoting overall well-being and
reducing health disparities in communities.

Education is a key social determinant
of health, influencing various aspects
of an individual's life. Higher levels
of education are associated with better
health outcomes, as education enhances
one's ability to make informed decisions
about lifestyle choices, healthcare
utilization, and disease prevention.
Limited educational opportunities can
result in lower health literacy, reducing
an individual's capacity to navigate
complex healthcare systems and understand
preventive measures. Educational
disparities contribute to a cycle of
poor health outcomes, as individuals
with lower levels of education may face
challenges in securing stable employment
and accessing adequate healthcare
resources. Addressing educational
inequalities is essential for fostering
a healthier and more equitable society.

Access to nutritious and sufficient
food is a critical social need that
directly impacts health. Food insecurity,
often linked to low income, can lead to
malnutrition, obesity, and various health
issues. Lack of access to fresh and
healthy foods contributes to the prevalence
of chronic conditions such as diabetes
and cardiovascular diseases. Current
issues related to food include disparities
in food availability, affordability, and
nutritional quality, disproportionately
affecting marginalized communities.
Addressing these issues involves promoting
food equity, supporting local agriculture,
and implementing policies that ensure
everyone has access to nutritious food,
ultimately improving overall public health.

Housing plays a vital role in determining
health outcomes. Inadequate or unstable
housing conditions can lead to physical
and mental health issues. Homelessness
and substandard living conditions expose
individuals to environmental hazards,
increase stress levels, and contribute
to the spread of infectious diseases.
The current housing crisis in many regions
exacerbates these issues, with rising
costs and limited affordable housing
options. Addressing housing as a social
need involves implementing policies to
ensure affordable housing, preventing
homelessness, and improving living
conditions to create a foundation for
better health outcomes.

Access to healthcare through insurance
is a critical social need that significantly
impacts health outcomes. Lack of health
insurance can lead to delayed or forgone
medical care, resulting in the progression
of illnesses and poorer health.
Disparities in insurance coverage contribute
to inequalities in healthcare access and
health outcomes. Current issues include
the high costs of healthcare and the
existence of uninsured or underinsured
populations. Addressing these challenges
involves expanding access to affordable
health insurance, implementing healthcare
reforms, and promoting policies that
ensure comprehensive coverage for all
individuals.

Rehospitalization refers to the phenomenon
where individuals experience repeated
hospital admissions, often due to
complications or inadequate follow-up care.
High rates of rehospitalization indicate
gaps in healthcare delivery and management,
impacting both individual well-being and
healthcare system efficiency. This social
need is indicative of issues such as
insufficient post-discharge support,
medication management, and coordination
of care. Repeated hospitalizations not
only strain healthcare resources but also
have significant implications for the
individuals involved, leading to increased
morbidity, reduced quality of life, and
higher healthcare costs. Addressing
rehospitalization requires improvements
in transitional care, enhanced coordination
among healthcare providers, and better
support systems for patients, especially
those with chronic conditions, to ensure
a smoother recovery and reduce the burden
on healthcare systems. Recognizing and
addressing these factors can contribute
to overall health improvement, prevent
unnecessary healthcare utilization, and
enhance the effectiveness of healthcare
services.

@@

- The data will be presented in Section\ref{sec:data}

- The methods are described in Section\ref{sec:meth}

- The results are reported in Section\ref{sec:resu}

- A discussion concludes in Section ref{sec:disc}

@@

In this study, descriptive statistics is utilized to outline the total population, racial composition, 
and education levels across the 8 counties in Connecticut. ANOVA analyses were 
conducted to assess the significance of the difference of the variables of median income, health insurance, food access, housing, and rehospitalization rates between counties and across the four years. Additional 
Tukey's HSD tests were conducted to determine the specific counties and years that have had significant differences within 
each of the variables for each county.

@@

County names are assigned to a respective county number: 
1	Fairfield County,
2	Hartford County,
3	Litchfield County,
4	Middlesex County,
5	New Haven County,
6	New London County,
7	Tolland County,
8	Windham County.

The racial and ethnic distribution of the population of the 8 counties investigated in this study is displayed in the following graphs:
\ref{tab:Demographics1}
\ref{tab:Demographics2}

@@

  \begin{figure}[tbp]
    \label{fig:Social Needs by YearGraph}
      \centering
      \includegraphics[width=\textwidth]{By Year.pdf}
      \caption{Social Needs Multiple Comparisons by County.}
    \end{figure}

@@

    Fairfield County & Hartford County & 20565.50* & 1883.47 & \textbf{\textless0.001} & 14327.60 & 26803.40 \\
    & Litchfield County & 14919.50* & 1883.47 & \textbf{\textless0.001} & 8681.60 & 21157.40 \\
    & Middlesex County & 9671.75* & 1883.47 & \textbf{\textless0.001} & 3433.85 & 15909.65 \\
    & New Haven County & 25662.75* & 1883.47 & \textbf{\textless0.001} & 19424.85 & 31900.65 \\
    & New London County & 21456.50* & 1883.47 & \textbf{\textless0.001} & 15218.60 & 27694.40 \\

@@

    & Windham County & 28671.00* & 1883.47 & \textbf{\textless0.001} & 22433.10 & 34908.90 \\

@@

    Hartford County & Fairfield County & -20565.50* & 1883.47 & \textbf{\textless0.001} & -26803.40 & -14327.60 \\

@@

    & Middlesex County & -10893.75* & 1883.47 & \textbf{\textless0.001} & -17131.65 & -4655.85 \\

@@

    & Tolland County & -11860.50* & 1883.47 & \textbf{\textless0.001} & -18098.40 & -5622.60 \\

@@

    Litchfield County & Fairfield County & -14919.50* & 1883.47 & \textbf{\textless0.001} & -21157.40 & -8681.60 \\

@@

    & New Haven County & 10743.25* & 1883.47 & \textbf{\textless0.001} & 4505.35 & 16981.15 \\

@@

    & Windham County & 13751.50* & 1883.47 & \textbf{\textless0.001} & 7513.60 & 19989.40 \\

@@

    Middlesex County & Fairfield County & -9671.75* & 1883.47 & \textbf{\textless0.001} & -15909.65 & -3433.85 \\
    & Hartford County & 10893.75* & 1883.47 & \textbf{\textless0.001} & 4655.85 & 17131.65 \\

@@

    & New Haven County & 15991.00* & 1883.47 & \textbf{\textless0.001} & 9753.10 & 22228.90 \\
    & New London County & 11784.75* & 1883.47 & \textbf{\textless0.001} & 5546.85 & 18022.65 \\

@@

    & Windham County & 18999.25* & 1883.47 & \textbf{\textless0.001} & 12761.35 & 25237.15 \\

@@

    New Haven County & Fairfield County & -25662.75* & 1883.47 & \textbf{\textless0.001} & -31900.65 & -19424.85 \\

@@

    & Litchfield County & -10743.25* & 1883.47 & \textbf{\textless0.001} & -16981.15 & -4505.35 \\
    & Middlesex County & -15991.00* & 1883.47 & \textbf{\textless0.001} & -22228.90 & -9753.10 \\

@@

    & Tolland County & -16957.75* & 1883.47 & \textbf{\textless0.001} & -23195.65 & -10719.85 \\

@@

    New London County & Fairfield County & -21456.50* & 1883.47 & \textbf{\textless0.001} & -27694.40 & -15218.60 \\

@@

    & Middlesex County & -11784.75* & 1883.47 & \textbf{\textless0.001} & -18022.65 & -5546.85 \\

@@

    & Tolland County & -12751.50* & 1883.47 & \textbf{\textless0.001} & -18989.40 & -6513.60 \\

@@

    & Hartford County & 11860.50* & 1883.47 & \textbf{\textless0.001} & 5622.60 & 18098.40 \\

@@

    & New Haven County & 16957.75* & 1883.47 & \textbf{\textless0.001} & 10719.85 & 23195.65 \\
    & New London County & 12751.50* & 1883.47 & \textbf{\textless0.001} & 6513.60 & 18989.40 \\
    & Windham County & 19966.00* & 1883.47 & \textbf{\textless0.001} & 13728.10 & 26203.90 \\

@@

    & Litchfield County & -13751.50* & 1883.47 & \textbf{\textless0.001} & -19989.40 & -7513.60 \\
    & Middlesex County & -18999.25* & 1883.47 & \textbf{\textless0.001} & -25237.15 & -12761.35 \\

@@

    & Tolland County & -19966.00* & 1883.47 & \textbf{\textless0.001} & -26203.90 & -13728.10 \\

@@

    * The mean difference is significant at the 0.05 level\\

@@

\begin{table}[]
  \label{tab: Multiple Comparisons}
  \caption{Multiple Comparisons by Year and Town}
  \resizebox{\columnwidth}{!}{%
  \begin{tabular}{llrrrrl}
  \textbf{Social Need}   & \textbf{Comparison by} & \multicolumn{1}{l}{\textbf{DF}} & \multicolumn{1}{l}{\textbf{Sum of Squares}} & \multicolumn{1}{l}{\textbf{Mean Square}} & \multicolumn{1}{l}{\textbf{F Value}} & \textbf{p Value}           \\
  Median Income          & Year                   & 3                               & 158657606                                   & 52885869                                 & 95.57                                & \textbf{\textless0.001}                    \\
                         & Town                   & 7                               & 2618521106                                  & 374074444                                & 675.99                               & \textbf{\textless0.001}                      \\
  Food                   & Year                   & 3                               & 3.6450625                                   & 1.2150208                                & 10.08                                & \textbf{0.0003} \\
                         & Town                   & 7                               & 348.3382875                                 & 49.7626125                               & 412.72                               & \textbf{\textless0.001}                      \\
  Housing                & Year                   & 3                               & 1.4535125                                   & 0.4845042                                & 0.41                                 & 0.7506 \\
                         & Town                   & 7                               & 183.9498375                                 & 26.2785482                               & 21.99                                & \textbf{\textless0.001}                      \\
  Rehospitalization Rate & Year                   & 3                               & 0.00050938                                  & 0.00016979                               & 1.45                                 & 0.2578 \\
                         & Town                   & 7                               & 0.00427188                                  & 0.00061027                               & 5.2                                  & \textbf{0.0015} \\
  Uninsured Rate         & Year                   & 3                               & 2.6848375                                   & 0.89494583                               & 17.17                                & \textbf{\textless0.001}                     \\
                         & Town                   & 7                               & 92.8554875                                  & 13.26506964                              & 254.57                               & \textbf{\textless0.001}    \\                
  * The mean difference is significant at the 0.05 level\\
  \end{tabular}%
  }
  \end{table}

@@

\ref{fig:Education by County Graph}
discuss, use lit to support why(descriptive) specific to CT

Table 4\ref{fig:Social Needs by County Graph} reveals that there is a signifiant difference between 
social need variables by year (2017-2020) for median income, food, and uninsured rate. Table 4 also 
reveals that all five variables analyzes(median income, food, housing, rehospitalization rate, and 
uninsured rate) are significantly difference. Multiple Comparison tests are conducted to identify the 
specific similarities and differences between the 8 CT counties.
\\
I. Comparison by County
\\
The results of the Tukey multiple comparison analysis\ref{tab: Multiple Comparisons} shed light on the relationship between county and key 
variables such as median income, housing affordability, rehospitalization rates, food stamps usage, and
uninsured rates. Understanding these correlations is vital in comprehending 
the socio-economic and racial disparities prevalent in the studied region.

The disparity in median income over the studied years, particularly the peak observed in 2020, 
signifies economic fluctuations. Notably, county 1 stands out with significantly higher income 
compared to other counties, indicating potential disparities in resource access and opportunities. 
\ref{tab:Multiple Comparison for Income} displays the individual significant levels for comparison with County 1.
All counties 2-8 appear to be significantly different at p <0.05 compared to county 1.
This wealth gap implies potential repercussions on community health outcomes. The consistent trend in 
the percentage of income spent on rent underscores a persistent challenge for residents, especially 
noticeable in counties 1 and 5, indicating enduring economic pressure.

@@

These analyses reveals the presence of a drastic inequality in County 1(Fairfield County). Although Farfield County
is revealed to have the highest median income, greatest educational attainment, Farfield County also has the greatest 
percent population uninsured with health insurance and the greatest proportion of population whose renters pay rent 
50 percent of their income. According to a study by the Federal Reserve Bank of New York{ ???????}, the most highest paid 
workers in Fairfield County earn a salary that is nine times greater than their lower-wage counterparts. Technology causing the disparities
\\
II. Comparison by Year
\\
An analysis spanning four years indicates significant differences in median income, with the highest observed 
in 2020 and the lowest in 2017. While there doesn't seem to be a notable overall difference in the proportion 
of the population relying on food stamp programs or facing uninsured rates, distinctions emerge when considering 
specific timeframes. The use of food stamps was higher in 2020 and 2019 compared to 2017, showcasing a 
shift in reliance on such programs over the years. Similarly, the uninsured rates were notably higher in 2020 and 
2018 compared to 2019. These findings suggest despite an overall increase in income over time, there is a simultaneous 
growing demand for food stamp programs and a pressing need for improved access to affordable health insurance. 
This suggests that, while income is on the rise, significant portions of the population still face challenges in
 meeting basic needs and accessing essential services. Addressing these trends requires comprehensive strategies 
 that go beyond income growth alone.
